\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-08-13 NAR_DRAFT.tex
commit to merge

Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all observed tumor mutations of substitution type $\mathrm{S}_{i,j}$.
The distribution of observed substitutions $ \mathrm{p}(\mathrm{S}) $, is multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{m,c}\} $.
In a tumor with a very large number of verified mutations $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile, but in practice we find that many elements of $\boldsymbol{\pi}$ will have $0$ or a very small number of mutations present in the tumor.
As a result we need to model $\mathrm{p}(\mathrm{S})$ as dirichlet - multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$,

@@

\begin{equation}
  \label{eqn:computed_posterior}
  $$
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{logit}_{10}(\mathrm{p}(\mathrm{S_{m,c}}))},
  $$
\end{equation}
where $\mathrm{p}(\mathrm{S_{i,j}})$ is the prior probability of a substitution of type $\mathrm{S_{m,c}}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
When comparing our posterior odds to those of MuTect in this paper, we use the assumption put forward by the MuTect authors in \citet{Cibulskis2013}.
They suggest a threshold for the posterior odds of 2, and a prior per-base probability of mutation of $3\cdot10^{-6}$, leading to a \texttt{TLOD} threshold of 6.3.
We compute the posterior odds for Mutect as
\begin{equation}
  \label{eqn:mutect_posterior}
  $$
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6},
  $$
\end{equation}
since $\mathrm{logit}_{10}(3\cdot10^{-6}) \approx -6$, and $10^{.3} \approx 2$.

>>>>

Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all observed tumor mutations of substitution type $\mathrm{S}_{m,c}$.
The distribution of observed substitutions $ \mathrm{p}(\mathrm{S}) $, is multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{m,c}\} $.
In a tumor with a very large number of verified mutations $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile, but in practice we find that many elements of $\boldsymbol{\pi}$ will have $0$ or a very small number of mutations present in the tumor.
As a result we need to model $\mathrm{p}(\mathrm{S})$ as dirichlet - multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$,

@@

\begin{equation}
  \label{eqn:computed_posterior}
  $$
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{logit}_{10}(\mathrm{p}(\mathrm{S_{m,c}}))},
  $$
\end{equation}
where $\mathrm{p}(\mathrm{S_{m,c}})$ is the prior probability of a substitution of type $\mathrm{S_{m,c}}$, as described in equation \ref{eqn:detailed_prior} and specified in equations \ref{eqn:post_pred} - \ref{eqn:mut_rate}.
When comparing our posterior odds to those of MuTect in this paper, we use the assumption put forward by the MuTect authors in \citet{Cibulskis2013}.
They suggest a threshold for the posterior odds of 2, and a prior per-base probability of mutation of $3\cdot10^{-6}$, leading to a \texttt{TLOD} threshold of 6.3.
We compute the posterior odds for Mutect as
\begin{equation}
  \label{eqn:mutect_posterior}
  $$
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6},
  $$
\end{equation}
since $\mathrm{logit}_{10}(3\cdot10^{-6}) \approx -6$, and $10^{.3} \approx 2$.
